Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-05-28 src/tex/ms.tex
[showyourwork] overleaf sync

\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\T}[1]{\textcolor{orange}{#1}}
%\newcommand{\U}[1]{\textbf{\textit{\textcolor{blue}{#1}}}}
\newcommand{\U}[1]{\textcolor{black}{#1}}
\usepackage{subfigure}

@@

\subsection{Bayesian model}
\texttt{CHEMPY} effectively is a Bayesian model for stellar abundances given a set of parameters $\{\Vec{\Lambda},\Vec{\Theta},T\}$ and \citet{Philcox_2019} extended the \texttt{CHEMPY} framework to be able to model multiple stellar data-points. Consider a given star with index $i$ that is born in some region of the ISM. This star will carry its own set of parameters $\{\Vec{\Lambda},\Vec{\Theta}_i,T_i\}$, where $\Vec{\Lambda}$ are star independent and hence taken to be global parameters while the ISM parameters $\Vec{\Theta}_i$ and the birth-time $T_i$ are star specific. Using \texttt{CHEMPY} (or the trained neural network emulator) we can easily model the set of $n_\mathrm{el}$ chemical abundances $\{X_i^j\}$ for the $i$-th star as: 
\begin{equation}\label{eq:chempy_function}
\{X_i^j\} = \texttt{CHEMPY}(\Vec{\Lambda},\Vec{\Theta}_i,T_i),
\end{equation}
where $j$ indexes the chemical element. 
These model abundances can then be compared against observations, with measured abundances $d_i^j$ and corresponding Gaussian errors $\sigma_{i,\mathrm{obs}}^j$ jointly denoted as $D_i=\{d_i^j,\sigma_{i,\mathrm{obs}}^j\}$.

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section. Inputs are sampled from the Gaussian priors shown in Tab. \ref{tab:priors}. Training takes $\sim10$ minutes on an Apple M1 Max.
We evaluate the accuracy of the NPE using $5\times10^3$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate eq.~\ref{eq:posterior}. We circumvent this by approximating each single star posterior by a Gaussian and fit for the parameters of mean and covariance. With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'^2}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

We accompany the APE analysis by showing the full posterior inference results for a single star in Fig.~\ref{fig:corner_plot}. This figure shows that the SBI approach is well able to infer correct parameters and their cross-correlations. In partciular, there is a strong correlation between the two global parameters as seen in the top left corner as well as for time (or stellar age) and all other five parameters as visible from the bottom row.

@@

@@

An exciting avenue for future research involves expanding the SBI method presented here to be able to cope with missing data points by switching to a transformer model architecture.
Also, with slight modifications, SBI could be used 
to directly infer empirical nucleosynthetic yields, which remain a major uncertainty in GCE models. In the current framework, \texttt{CHEMPY} relies on tabulated yields from theoretical studies, which may not fully represent the complex processes driving stellar enrichment. Adapting SBI to simultaneously infer galactic parameters and refine empirical yield tables would require changes to the simulator. Specifically, \texttt{CHEMPY} would need to incorporate parameterized yield modifications as part of its input space, allowing for flexible adjustments to enrichment rates during inference. This would also necessitate larger training datasets and enhanced validation techniques to ensure convergence. Such an approach could provide a unified framework for calibrating galactic models directly against observational data. Research in this direction is currently performed and will be part of a future paper.

@@

The applicability of our method extends to surveys with more observed elements especially neutron capture elements such as GALAH DR4 \citep{Buder2024} and future spectroscopic surveys, such as those planned by the 4MOST \citep{4most} or WEAVE \citep{weave} consortia, which will provide orders of magnitude more data than current datasets. Our findings indicate that SBI can seamlessly scale to such large datasets, offering significant advantages in terms of speed and computational cost.

@@

@@

>>>>

\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\T}[1]{\textcolor{orange}{#1}}
%\newcommand{\U}[1]{\textbf{\textit{\textcolor{blue}{#1}}}}
\newcommand{\U}[1]{\textcolor{black}{#1}}
\newcommand{\resub}[1]{\textbf{\textcolor{blue}{#1}}}
\usepackage{subfigure}

@@

\subsection{Bayesian model}
\label{subsec:bayesian-model}
\texttt{CHEMPY} effectively is a Bayesian model for stellar abundances given a set of parameters $\{\Vec{\Lambda},\Vec{\Theta},T\}$ and \citet{Philcox_2019} extended the \texttt{CHEMPY} framework to be able to model multiple stellar data-points. Consider a given star with index $i$ that is born in some region of the ISM. This star will carry its own set of parameters $\{\Vec{\Lambda},\Vec{\Theta}_i,T_i\}$, where $\Vec{\Lambda}$ are star independent and hence taken to be global parameters while the ISM parameters $\Vec{\Theta}_i$ and the birth-time $T_i$ are star specific. Using \texttt{CHEMPY} (or the trained neural network emulator) we can easily model the set of $n_\mathrm{el}$ chemical abundances $\{X_i^j\}$ for the $i$-th star as: 
\begin{equation}\label{eq:chempy_function}
\{X_i^j\} = \texttt{CHEMPY}(\Vec{\Lambda},\Vec{\Theta}_i,T_i),
\end{equation}
where $j$ indexes the chemical element. 
These model abundances can then be compared against observations, with measured abundances $d_i^j$ and corresponding Gaussian errors $\sigma_{i,\mathrm{obs}}^j$ jointly denoted as $D_i=\{d_i^j,\sigma_{i,\mathrm{obs}}^j\}$.

@@

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section. Inputs are sampled from the Gaussian priors shown in Tab. \ref{tab:priors}. Training takes $\sim10$ minutes on an Apple M1 Max.
We evaluate the accuracy of the NPE using $5\times10^3$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, \resub{we add 5\%  Gaussian observational uncertainties ($=0.05$ dex) to the abundances simulated with \texttt{CHEMPY}, representative of typical APOGEE measurements \citep{apogee17}, to simulate observational noise. While this uniform treatment simplifies the modeling, we acknowledge that in real data, uncertainties can vary by stellar type, element, or SNR. Future extensions of this work will incorporate more realistic, element-specific or SNR-dependent noise models.}

We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

This does however present one issue, that since we only have access to samples from the posterior and not the posterior itself it is difficult to evaluate eq.~\ref{eq:posterior}. \resub{We circumvent this by approximating each single-star posterior by a Gaussian and fitting for the parameters of mean and covariance. This choice enables an analytical combination of multiple posteriors (see Eq.~\ref{eq:posterior}), providing computational tractability and allowing efficient inference over large datasets. While this approximation may limit the flexibility of the posterior, it significantly reduces computational complexity. Alternative strategies, such as using a Mixture Density Network (MDN) or normalizing flows with parametric forms, would provide richer posterior representations. However, these would complicate the composition of posteriors across stars. We acknowledge this as a simplification and plan to explore more expressive architectures in future work.}

With this it is straight forward to evaluate eq.~\ref{eq:posterior} analytically. In fact, under our assumption the combined posterior is a product between the prior and the product of multiple Gaussians for the individual star posteriors. The latter product is also a Gaussian with mean $\mathbf{\mu'}$ and variance $\mathbf{\sigma'^2}$:
\begin{align}
    \mathbf{\mu'} &= \frac{\sum_{i=1}^{N_{stars}} \frac{\mu_i}{\sigma_i^2}}{\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}} \\
\mathbf{\sigma'}^2 &= \frac1 {\sum_{i=1}^{N_{stars}} \frac1{\sigma_i^2}}
\end{align}

@@

We accompany the APE analysis by showing the full \resub{joint posterior over global and local parameters for a single star in Fig.~\ref{fig:corner_plot}. When performing inference across multiple stars, we marginalize over local parameters and use the factorized form in Eq. 5 to approximate the global posterior. This distinction between single-star and population-level inference will be emphasized in the revised captions and text for clarity.}
This figure shows that the SBI approach is well able to infer correct parameters and their cross-correlations. In partciular, there is a strong correlation between the two global parameters as seen in the top left corner as well as for time (or stellar age) and all other five parameters as visible from the bottom row.

@@

\resub{It is important to note that our SBI results are derived using 1,000 stars, whereas the HMC results from \citet{Philcox_2019} are based on only 200 stars and involve slightly different HMC setups with additional parameters reflecting the individual element abundance weights or model errors contributing to the inference \citep[see section 4 of][]{Philcox_2019}. This contributes to the observed differences in posterior contours. Additionally, our method approximates posteriors using Gaussian fits, which may underestimate uncertainty at large sample sizes (see also Sec.~\ref{sec: discussion}). To ensure a fairer comparison, in appendix~\ref{sec:hmc-sbi} we have run a comparison between our SBI model and a custom HMC model reflecting the exact same setup as for the SBI. Figure~\ref{fig:} in the appendix shows that ...}

@@

An exciting avenue for future research involves expanding the SBI method presented here to be able to cope with missing data points by switching to a transformer model architecture.
\resub{A natural question arises regarding the use of these alternative architectures -- such as transformer-based models -- which could process variable numbers of stellar inputs directly, bypassing the need for single-star posterior approximations. These models could potentially capture correlations across stars and remove the need for assuming a factorized posterior. However, incorporating such models would require a new approach to network training and posterior inference. We are currently investigating these strategies and will report results in a forthcoming paper.}

Also, with slight modifications, SBI could be used 
to directly infer empirical nucleosynthetic yields, which remain a major uncertainty in GCE models. In the current framework, \texttt{CHEMPY} relies on tabulated yields from theoretical studies, which may not fully represent the complex processes driving stellar enrichment. Adapting SBI to simultaneously infer galactic parameters and refine empirical yield tables would require changes to the simulator. Specifically, \texttt{CHEMPY} would need to incorporate parameterized yield modifications as part of its input space, allowing for flexible adjustments to enrichment rates during inference. This would also necessitate larger training datasets and enhanced validation techniques to ensure convergence. Such an approach could provide a unified framework for calibrating galactic models directly against observational data. Research in this direction is currently performed and will be part of a future paper.

@@

The applicability of our method extends to surveys with more observed elements especially neutron capture elements such as GALAH DR4 \citep{Buder2024} and future spectroscopic surveys, such as those planned by the 4MOST \citep{4most} or WEAVE \citep{weave} consortia, which will provide orders of magnitude more data than current datasets. Our findings indicate that SBI can seamlessly scale to such large datasets, offering significant advantages in terms of speed and computational cost.
\resub{While the present study focuses on mock datasets to validate the methodology under controlled conditions, we recognize the importance of testing our method on real survey data. Ongoing work is aimed at applying our pipeline to GALAH DR4 and APOGEE datasets. These surveys offer diverse chemical tracers and provide a valuable testbed for evaluating the robustness and calibration of our approach on real observations. This will be presented in future work.}

@@

\resub{The calibration tests shown here are performed using posterior samples from our NPE trained on CHEMPY mock data. Validation is carried out on a hold-out set of simulated stars, with Gaussian noise added (0.05 dex), following the same protocol as described in Sec.~\ref{subsec:bayesian-model}. These tests evaluate the quality of posterior estimates for the global parameters  using single-star inference.}

@@

\section{\resub{Comparison between HMC and SBI}}
\label{sec:hmc-sbi}
\resub{put a new figure and some text...}

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/hmc_corner_plot.pdf}
    \vspace{-.5cm}
    \caption{Caption}
    \label{fig:hmc}
    \script{hmc.py}
\end{figure}
